\newpage
\chapter*{\centering Abstract}

Managing recruitment for institutional projects in educational institutions often involves manual handling of applications through emails or paper documents, leading to fragmented applicant data, difficulty in managing multiple projects, and limited transparency. The Institutional Recruitment Portal is a secure, web-based platform designed to streamline and digitize the end-to-end recruitment workflow for project-specific hiring. The system provides a unified environment for Administrators, Professors, and Applicants.

The platform enables professors to create and manage project listings, set deadlines, and define required documents. Applicants can securely register, apply for specific projects, and upload necessary credentials such as ID proofs, academic documents, and payment receipts. The system implements a structured local filesystem storage architecture, replacing traditional cloud dependencies for better institutional control and faster document management. Security is ensured through JSON Web Token (JWT) authentication, Role-Based Access Control (RBAC), and bcrypt password hashing.

The system is built on the MERN stack (MongoDB, Express.js, React, Node.js), containerized with Docker, and served through Nginx. Key administrative features include professor management, project moderation, and comprehensive application tracking. The portal also supports bulk document downloads as ZIP archives and application data export in CSV format, facilitating efficient post-application processing.

Evaluation demonstrates that the Institutional Recruitment Portal significantly reduces manual overhead, improves data accessibility for professors, and provides a professional, streamlined experience for applicants, thereby establishing a robust digital infrastructure for institutional project management.
